---
title: "Join Lab"
author: "Bryce Borak"
date: "DATA 2401"
output: html_document
---


# Joins and `dplyr` practice

## Dataset 1

We start by making some "healthcare" data: 

```{r}
# Patients table
patients <- tibble(
  patient_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 40, 35, 50),
  sex = c("F", "M", "F", "M")
)

# Visits table (Some patients have multiple visits, some missing)
visits <- tibble(
  visit_id = c(101, 102, 103, 104, 105),
  patient_id = c(1, 2, 2, 3, 5),  
  doctor_id = c(201, 202, 201, 203, 204),
  visit_date = as.Date(c("2024-01-10", "2024-01-12", "2024-02-15", "2024-03-01", "2024-03-10"))
)

# Doctors table (Some doctors may not have treated any patients)
doctors <- tibble(
  doctor_id = c(201, 202, 203, 205),
  doctor_name = c("Dr. Smith", "Dr. Johnson", "Dr. Lee", "Dr. Brown"),
  specialty = c("Cardiology", "Neurology", "Pediatrics", "Orthopedics")
)

```

Questions: 

- Find the names of patients who have visited the hospital. 
- Show all patients visit information, even if they haven't visited the hospital
- List all the visits and patient information, even if there isn't information for the patient
- Show all visits with their assigned doctors.
- Find the doctor(s) that Bob has met with. 
- Find patients that haven't visited the hospital

```{r}
library(tidyverse)

# 1) Find the names of patients who have visited the hospital.
patients %>%
  inner_join(visits, by = "patient_id") %>%
  distinct(name)

# 2) Show all patients visit information, even if they haven't visited the hospital
patients %>%
  left_join(visits, by = "patient_id") %>%
  arrange(patient_id, visit_date)

# 3) List all the visits and patient information, even if there isn't information for the patient
visits %>%
  left_join(patients, by = "patient_id") %>%
  arrange(visit_id)

# 4) Show all visits with their assigned doctors.
visits %>%
  left_join(doctors, by = "doctor_id") %>%
  arrange(visit_id)

# 5) Find the doctor(s) that Bob has met with.
patients %>%
  filter(name == "Bob") %>%
  inner_join(visits, by = "patient_id") %>%
  inner_join(doctors, by = "doctor_id") %>%
  distinct(doctor_name, specialty)

# 6) Find patients that haven't visited the hospital
patients %>%
  anti_join(visits, by = "patient_id") %>%
  select(patient_id, name, age, sex)
```

## Dataset 2

Load the `"nycflights13"` package.  You'll also need to load `dplyr` or `tidyverse`, if you didn't above.

```{r}
library(tidyverse)
#install.packages("nycflights13")
library(nycflights13)
my_flights <- flights # store my own version
my_flights <- na.omit(my_flights) ## this dataset has data on cancelled flights too, 
                                  ## so we 'll remove those for now

```


Create a dataframe of the average arrival delays for each _destination_

```{r}
avg_arr_by_dest <- my_flights %>%
  group_by(dest) %>%
  summarise(
    n_flights = n(),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  )

avg_arr_by_dest
```

Then use `left_join()` to join on the "airports" dataframe, which has the airport information.

```{r}
avg_arr_by_dest_airport <- avg_arr_by_dest %>%
  left_join(airports, by = c("dest" = "faa"))

avg_arr_by_dest_airport
```


Note here: the column names don't match anymore! 

Which airport had the largest average arrival delay?

```{r}
avg_arr_by_dest_airport %>%
  arrange(desc(avg_arr_delay)) %>%
  select(dest, name, n_flights, avg_arr_delay) %>%
  slice(1)
```



What is the average departure delay for each airline? To answer this with something more than just the FAA code of the airline name, you need to create a dataframe of the average departure delay for each _airline_, then use `left_join()` to join on the "airlines" dataframe


```{r}
avg_dep_by_carrier <- my_flights %>%
  group_by(carrier) %>%
  summarise(
    n_flights = n(),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(airlines, by = "carrier") %>%
  arrange(avg_dep_delay)

avg_dep_by_carrier

```

Which airline had the smallest average departure delay? 



```{r}
avg_dep_by_carrier %>%
  select(carrier, name, n_flights, avg_dep_delay) %>%
  slice(1)


```


Add origin and destination airport names (two joins to the same table)
Question: Which origin–destination pair has the highest average arrival delay?


```{r}
od_avg_arr <- my_flights %>%
  group_by(origin, dest) %>%
  summarise(
    n_flights = n(),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(airports %>% select(faa, origin_name = name),
            by = c("origin" = "faa")) %>%
  left_join(airports %>% select(faa, dest_name = name),
            by = c("dest" = "faa"))

od_avg_arr %>%
  arrange(desc(avg_arr_delay)) %>%
  select(origin, origin_name, dest, dest_name, n_flights, avg_arr_delay) %>%
  slice(1)


``` 

Which airline (name) has the highest on-time rate? Are left_join and inner_join returning the same rows here? Why?


```{r}
on_time_by_carrier <- my_flights %>%
  mutate(on_time = arr_delay <= 0) %>%
  group_by(carrier) %>%
  summarise(
    n_flights = n(),
    on_time_rate = mean(on_time, na.rm = TRUE),
    .groups = "drop"
  )

on_time_left  <- on_time_by_carrier %>% left_join(airlines, by = "carrier")
on_time_inner <- on_time_by_carrier %>% inner_join(airlines, by = "carrier")

# Highest on-time rate
on_time_left %>%
  arrange(desc(on_time_rate)) %>%
  select(carrier, name, n_flights, on_time_rate) %>%
  slice(1)

# Are left_join and inner_join the same rows?
nrow(on_time_left)
nrow(on_time_inner)
anti_join(on_time_left, on_time_inner, by = c("carrier", "n_flights", "on_time_rate", "name"))


```

Plane metadata join: age of aircraft vs. delays
Among aircraft with at least 50 flights, which tailnumber has the largest average arrival delay, and what are the manufacturer/model and age?


```{r}
tail_delay <- my_flights %>%
  filter(!is.na(tailnum)) %>%
  group_by(tailnum) %>%
  summarise(
    n_flights = n(),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_flights >= 50) %>%
  left_join(planes, by = "tailnum") %>%
  mutate(age_2013 = 2013 - year)

tail_delay %>%
  arrange(desc(avg_arr_delay)) %>%
  select(tailnum, n_flights, avg_arr_delay, manufacturer, model, year, age_2013) %>%
  slice(1)


```

## Extra Examples from R4DS:    

```{r}

top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest

```

Now you want to find each flight that went to one of those destinations. You could, of course, filter. But! Now you can semi-join: 

```{r}
flights %>% 
  semi_join(top_dest)
```


What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?
Perform the two joins below and consider what the results **are**. 

```{r}
# Flights whose dest code is NOT in airports$faa (destinations missing from airports)
anti_join(flights, airports, by = c("dest" = "faa"))

# Airports that never show up as a destination in flights$dest
anti_join(airports, flights, by = c("faa" = "dest"))
```
